A Bioconductor workflow for processing and analysing spatial proteomics data

### 2018-04-17 bioc-workflow-resubmit.Rnw
don't load and attach gridExtra
fix img label
small changes to intro

Quantitative mass spectrometry based spatial proteomics involves
elaborate, expensive and time consuming experimental protocols and
considerable effort is invested in the generation of such data.
Multiple research groups have described a variety of approaches to
establish high quality proteome-wide datasets (see for example
\cite{Gatto:2010} for a review, and
\cite{hyper,Itzhak:2016,Jean_Beltran:2016} for recent
examples). However, data analysis is as critical as data production
for reliable and insightful biological interpretation. Here, we walk
the reader through a typical pipeline for the analysis of such data
using several Bioconductor~\cite{Huber:2015} packages for the R
statistical programming environment.

@@

Installation of Bioconductor package is documented in detail on the
\href{http://bioconductor.org/install/#install-bioconductor-packages}{Bioconductor
  installation help page}. Below, we show how to install the four main
packages used in this workflow:

@@

\begin{figure}[!ht]
  \centering
<<qcbx, out.width="\\textwidth", fig.asp=1/2>>=
par(mfrow = c(1, 2), ## creates a two-panel figure
    las = 2,         ## axis labels orientation
    cex.axis = .7)   ## axis label size
o <- order(hl$Iodixonal.Density)
plotDist(hl[, o], pcol = "#00000010", xlab = "")
lines(colMeans(exprs(hl[, o])), col = "red", type = "b")
boxplot(exprs(hl[, o]))
@  
\caption{Protein profiles and distribution of channel intensities. The
  red dots represent the mean relative intensity for each channel. }
  \label{fig:q  cbx}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
<<visualiseOptHide, echo=FALSE, out.width=".65\\textwidth">>=
library('gridExtra')
grid.arrange(plot(params), levelPlot(params), ncol = 2)
@  
  \caption{Assessment of the classification model parameter optimisation.}
  \label{fig:visualisOptHide}
\end{figure}

>>>>

Quantitative mass spectrometry-based spatial proteomics involves
elaborate, expensive and time consuming experimental protocols and
considerable effort is invested in the generation of such data.
Multiple research groups have described a variety of approaches to
establish high quality proteome-wide datasets (see for example
\cite{Gatto:2010} for a review, and
\cite{hyper,Itzhak:2016,Jean_Beltran:2016,Itzhak:2017,Hirst:2018} for
recent examples). However, data analysis is as critical as data
production for reliable and insightful biological
interpretation. Here, we walk the reader through a typical pipeline
for the analysis of such data using several
Bioconductor~\cite{Huber:2015} packages for the R statistical
programming environment.

@@

Installation of Bioconductor packages is documented in detail on the
\href{http://bioconductor.org/install/#install-bioconductor-packages}{Bioconductor
  installation help page}. Below, we show how to install the four main
packages used in this workflow:

@@

\begin{figure}[!ht]
  \centering
<<qcbx, out.width="\\textwidth", fig.asp=1/2>>=
par(mfrow = c(1, 2), ## creates a two-panel figure
    las = 2,         ## axis labels orientation
    cex.axis = .7)   ## axis label size
o <- order(hl$Iodixonal.Density)
plotDist(hl[, o], pcol = "#00000010", xlab = "")
lines(colMeans(exprs(hl[, o])), col = "red", type = "b")
boxplot(exprs(hl[, o]))
@
\caption{Protein profiles and distribution of channel intensities. The
  red dots represent the mean relative intensity for each channel. }
  \label{fig:qcbx}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
<<visualiseOptHide, echo=FALSE, out.width=".65\\textwidth">>=
gridExtra::grid.arrange(plot(params), levelPlot(params), ncol = 2)
@
  \caption{Assessment of the classification model parameter optimisation.}
  \label{fig:visualisOptHide}
\end{figure}

### 2018-04-17 bioc-workflow-resubmit.tex
small changes to intro

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{combined} \hlkwb{<-} \hlstd{MSnbase}\hlopt{::}\hlkwd{combine}\hlstd{(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)}
\hlstd{combined}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 6725 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2Y3-3 (6725 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Mon Apr 16 15:33:11 2018 
##  MSnbase version: 1.21.7
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{combined} \hlkwb{<-} \hlkwd{filterNA}\hlstd{(combined)}
\hlstd{combined}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Mon Apr 16 15:33:11 2018 
## Subset [6725,20][5032,20] Mon Apr 16 15:33:12 2018 
## Removed features with more than 0 NAs: Mon Apr 16 15:33:12 2018 
## Dropped featureData's levels Mon Apr 16 15:33:12 2018 
##  MSnbase version: 1.21.7
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{figure}[!ht]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{prots} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"Q8CG48"}\hlstd{,} \hlstr{"Q8CG47"}\hlstd{,} \hlstr{"Q8K2Z4"}\hlstd{,} \hlstr{"Q8C156"}\hlstd{)}
\hlstd{foi13s} \hlkwb{<-} \hlkwd{FeaturesOfInterest}\hlstd{(}\hlkwc{description} \hlstd{=} \hlstr{"13S consensin proteins"}\hlstd{,}
                             \hlkwc{fnames} \hlstd{= prots,}
                             \hlkwc{object} \hlstd{= hl)}
\hlstd{foi13s}
\end{alltt}
\begin{verbatim}
## Traceable object of class "FeaturesOfInterest"
##  Created on Mon Apr 16 15:33:14 2018 
##  Description:
##   13S consensin proteins
##  4 features of interest:
##    Q8CG48, Q8CG47, Q8K2Z4, Q8C156
\end{verbatim}
\begin{alltt}
\hlkwd{plot2D}\hlstd{(hl)}
\hlkwd{addLegend}\hlstd{(hl,} \hlkwc{cex} \hlstd{=} \hlnum{.6}\hlstd{)}
\hlkwd{highlightOnPlot}\hlstd{(hl, foi13s)}
\end{alltt}
\end{kframe}
\includegraphics[width=.8\textwidth]{figure/foi-1}

@@

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Mon Apr 16 15:33:14 2018
\begin{table}[ht]
\centering
\begin{tabular}{rrllr}
  \hline
 & Replicate & Tag & Gradient.Fraction & Iodixonal.Density \\ 
  \hline
X127C &   1 & 127C & 8 to 9 (pooled) & 11.00 \\ 
  X127C.1 &   2 & 127C & 7 to 9 (pooled) & 10.00 \\ 
   \hline
\end{tabular}
\caption{Differences in gradient fraction pooling.} 
\label{tab:pdtab}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{hl}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126 X127N ... X131.1 (20 total)
##   varLabels: Replicate Tag Gradient.Fraction Iodixonal.Density
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: uniprot.accession uniprot.id ... pd (27 total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Added markers from  'mrk' marker vector. Mon Apr 16 15:33:13 2018 
## Added markers from  'pdres' marker vector. Mon Apr 16 15:33:14 2018 
##  MSnbase version: 2.4.2
\end{verbatim}
\begin{alltt}
\hlkwd{getMarkers}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"pd"}\hlstd{)}
\end{alltt}
\begin{verbatim}
## organelleMarkers
##                          40S Ribosome 
##                                   106 
##                          60S Ribosome 
##                                    95 
## Endoplasmic reticulum/Golgi apparatus 
##                                   393 
##                         Mitochondrion 
##                                   525 
##                           Phenotype 1 
##                                   300 
##                           Phenotype 2 
##                                   253 
##                           Phenotype 3 
##                                   203 
##                           Phenotype 4 
##                                    74 
##                           Phenotype 5 
##                                    91 
##                       Plasma membrane 
##                                   421 
##                            Proteasome 
##                                    92 
##                               unknown 
##                                  2479
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{(ts} \hlkwb{<-} \hlkwd{orgQuants}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"svm"}\hlstd{,} \hlkwc{scol} \hlstd{=} \hlstr{"svm.scores"}\hlstd{,} \hlkwc{mcol} \hlstd{=} \hlstr{"markers"}\hlstd{,} \hlkwc{t} \hlstd{=} \hlnum{.5}\hlstd{))}
\end{alltt}
\begin{verbatim}
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4371207               0.3057074               0.3881852 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6925752               0.6021888               0.4265610 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4262321               0.5923175               0.9498363 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7961720               0.7109965               0.3179536 
##         Plasma membrane              Proteasome 
##               0.7201960               0.4170372
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4371207               0.3057074               0.3881852 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6925752               0.6021888               0.4265610 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4262321               0.5923175               0.9498363 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7961720               0.7109965               0.3179536 
##         Plasma membrane              Proteasome 
##               0.7201960               0.4170372
\end{verbatim}
\begin{alltt}
\hlstd{hl} \hlkwb{<-} \hlkwd{getPredictions}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"svm"}\hlstd{,} \hlkwc{scol} \hlstd{=} \hlstr{"svm.scores"}\hlstd{,} \hlkwc{mcol} \hlstd{=} \hlstr{"markers"}\hlstd{,} \hlkwc{t} \hlstd{= ts)}
\end{alltt}
\begin{verbatim}
## ans
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##                      84                     171                      89 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##                     294                     475                      97 
##    Extracellular matrix                Lysosome           Mitochondrion 
##                      29                     124                     523 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##                     229                     343                      39 
##         Plasma membrane              Proteasome                 unknown 
##                     317                     161                    2057
\end{verbatim}
\end{kframe}
\end{knitrout}

>>>>

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{combined} \hlkwb{<-} \hlstd{MSnbase}\hlopt{::}\hlkwd{combine}\hlstd{(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)}
\hlstd{combined}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 6725 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2Y3-3 (6725 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Tue Apr 17 08:39:23 2018 
##  MSnbase version: 1.21.7
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{combined} \hlkwb{<-} \hlkwd{filterNA}\hlstd{(combined)}
\hlstd{combined}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Tue Apr 17 08:39:23 2018 
## Subset [6725,20][5032,20] Tue Apr 17 08:39:24 2018 
## Removed features with more than 0 NAs: Tue Apr 17 08:39:24 2018 
## Dropped featureData's levels Tue Apr 17 08:39:24 2018 
##  MSnbase version: 1.21.7
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{figure}[!ht]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{prots} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"Q8CG48"}\hlstd{,} \hlstr{"Q8CG47"}\hlstd{,} \hlstr{"Q8K2Z4"}\hlstd{,} \hlstr{"Q8C156"}\hlstd{)}
\hlstd{foi13s} \hlkwb{<-} \hlkwd{FeaturesOfInterest}\hlstd{(}\hlkwc{description} \hlstd{=} \hlstr{"13S consensin proteins"}\hlstd{,}
                             \hlkwc{fnames} \hlstd{= prots,}
                             \hlkwc{object} \hlstd{= hl)}
\hlstd{foi13s}
\end{alltt}
\begin{verbatim}
## Traceable object of class "FeaturesOfInterest"
##  Created on Tue Apr 17 08:39:26 2018 
##  Description:
##   13S consensin proteins
##  4 features of interest:
##    Q8CG48, Q8CG47, Q8K2Z4, Q8C156
\end{verbatim}
\begin{alltt}
\hlkwd{plot2D}\hlstd{(hl)}
\hlkwd{addLegend}\hlstd{(hl,} \hlkwc{cex} \hlstd{=} \hlnum{.6}\hlstd{)}
\hlkwd{highlightOnPlot}\hlstd{(hl, foi13s)}
\end{alltt}
\end{kframe}
\includegraphics[width=.8\textwidth]{figure/foi-1}

@@

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Tue Apr 17 08:39:26 2018
\begin{table}[ht]
\centering
\begin{tabular}{rrllr}
  \hline
 & Replicate & Tag & Gradient.Fraction & Iodixonal.Density \\ 
  \hline
X127C &   1 & 127C & 8 to 9 (pooled) & 11.00 \\ 
  X127C.1 &   2 & 127C & 7 to 9 (pooled) & 10.00 \\ 
   \hline
\end{tabular}
\caption{Differences in gradient fraction pooling.} 
\label{tab:pdtab}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{hl}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126 X127N ... X131.1 (20 total)
##   varLabels: Replicate Tag Gradient.Fraction Iodixonal.Density
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: uniprot.accession uniprot.id ... pd (27 total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Added markers from  'mrk' marker vector. Tue Apr 17 08:39:25 2018 
## Added markers from  'pdres' marker vector. Tue Apr 17 08:39:26 2018 
##  MSnbase version: 2.4.2
\end{verbatim}
\begin{alltt}
\hlkwd{getMarkers}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"pd"}\hlstd{)}
\end{alltt}
\begin{verbatim}
## organelleMarkers
##                          40S Ribosome 
##                                   106 
##                          60S Ribosome 
##                                    95 
## Endoplasmic reticulum/Golgi apparatus 
##                                   393 
##                         Mitochondrion 
##                                   525 
##                           Phenotype 1 
##                                   300 
##                           Phenotype 2 
##                                   253 
##                           Phenotype 3 
##                                   203 
##                           Phenotype 4 
##                                    74 
##                           Phenotype 5 
##                                    91 
##                       Plasma membrane 
##                                   421 
##                            Proteasome 
##                                    92 
##                               unknown 
##                                  2479
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{(ts} \hlkwb{<-} \hlkwd{orgQuants}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"svm"}\hlstd{,} \hlkwc{scol} \hlstd{=} \hlstr{"svm.scores"}\hlstd{,} \hlkwc{mcol} \hlstd{=} \hlstr{"markers"}\hlstd{,} \hlkwc{t} \hlstd{=} \hlnum{.5}\hlstd{))}
\end{alltt}
\begin{verbatim}
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4363460               0.3023446               0.3833708 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6846500               0.5947855               0.4246911 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4302963               0.5753825               0.9498381 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7947918               0.7075039               0.3176585 
##         Plasma membrane              Proteasome 
##               0.7124225               0.4149409
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4363460               0.3023446               0.3833708 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6846500               0.5947855               0.4246911 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4302963               0.5753825               0.9498381 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7947918               0.7075039               0.3176585 
##         Plasma membrane              Proteasome 
##               0.7124225               0.4149409
\end{verbatim}
\begin{alltt}
\hlstd{hl} \hlkwb{<-} \hlkwd{getPredictions}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"svm"}\hlstd{,} \hlkwc{scol} \hlstd{=} \hlstr{"svm.scores"}\hlstd{,} \hlkwc{mcol} \hlstd{=} \hlstr{"markers"}\hlstd{,} \hlkwc{t} \hlstd{= ts)}
\end{alltt}
\begin{verbatim}
## ans
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##                      86                     170                      84 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##                     296                     474                      95 
##    Extracellular matrix                Lysosome           Mitochondrion 
##                      28                     123                     523 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##                     230                     344                      39 
##         Plasma membrane              Proteasome                 unknown 
##                     326                     157                    2057
\end{verbatim}
\end{kframe}
\end{knitrout}
